100 years of Pies vs. Bars

### 2025-06-28 Draft.qmd
added SS and edited until modern graphics

The debate between pie charts and bar charts has been ongoing for over a century.
Numerous studies with distinct methodologies and experimental designs have attempted to resolve this issue, but no single consensus has emerged.
In this paper, we will provide a comprehensive summary of the existing literature, analyze the strengths and limitations of both chart types, and attempt to identify a unified solution based on empirical evidence and theoretical insights.

@@

William Playfair introduced the pie and bar charts in the early 1800s, and by the turn of the next century, strong opinions about the effectiveness of alternative displays of the same data had been publicized in textbooks and manuals.
For instance, @eellsRelativeMeritsCircles1926 quotes two contemporary resources that discuss pie charts as "not a desirable form of presentation" [@brintonGraphicMethodsPresenting1914] and, more strongly, "an insult to a man's intelligence" [@karsten1923charts].
@brintonGraphicMethodsPresenting1914 even identifies that the "sector method" is common, but if horizontal bar methods were used more frequently, they would be preferred for both speed and accuracy.
@eellsRelativeMeritsCircles1926 also quotes an excerpt from an introductory statistics textbook [@secristIntroductionStatisticalMethods1925], which argues that pie charts require more cognitive effort to use accurately than a comparable stacked bar chart, calling pie diagrams "clumsy and defective".
These insights mirror today's status quo: pie charts are stubbornly common, even though many design guidelines recommend against their use

@wickhamGraphicalCriticismHistorical2013 previously noted the prescience of design guidelines prior to empirical studies validating their recommendations across different types of visualizations.

@@

Interestingly, the question of pies and bars motivated some of the first empirical evaluations of charts, published in the Journal of the American Statistical Association: @eellsRelativeMeritsCircles1926, @vonhuhnFurtherStudiesGraphic1927, and @croxtonBarChartsCircle1927.
Experimenters approached the problem head-on, directly asking participants to estimate quantities from charts, and evaluated participants on accuracy and, in some cases, speed.
While more sophisticated testing methodologies have been developed over the past century [@vanderplasTestingStatisticalCharts2020], studies are still conducted today using direct estimation approaches [@vanderplasFramedReproducingRevisiting2019].
The phrasing of direct estimation questions is of primary importance: a simple mathematical transformation of the estimated quantity can have an outsized effect on the results.

The initial studies of the question of pies and bars failed to identify a clear "best chart" across all situations or even to clearly identify in which situations pie charts or bar charts should be preferred.
@eellsRelativeMeritsCircles1926 found that pie charts were read as accurately as bar charts and with comparable speed.
The simplified nature of the experiment attracted critiques from @vonhuhnFurtherStudiesGraphic1927; attached to these criticisms is a short study previously run by Croxton, which examined estimates of the ratio between smaller and larger sectors of two-category pie and bar charts; these contributions will be addressed separately even though they are often cited as a single paper, even within the JASA records.
@vonhuhnFurtherStudiesGraphic1927's paper provides several examples which highlight concerns with the Eells study, and argues that while pie charts may be better in certain simple situations, bar charts are much more extensible and facilitate more advanced comparisons, but does not include any experimental studies.
<!-- Not much has changed today - experimental work is often criticized by those who provide toy examples but no empirical data due to compromises necessary to ensure experimental control.  --> Croxton's early experimental data (attached to @vonhuhnFurtherStudiesGraphic1927) had a larger number of participants than @eellsRelativeMeritsCircles1926, but only tested two-category charts with only two proportions: .25 and .4.
In addition, Croxton's study elicited estimates from participants as $A/B$ comparisons, where $A + B = 1$, while Eells elicited estimates for $A$ and $B$ separately.
<!-- It seems that prior to @eellsRelativeMeritsCircles1926, some researchers were empirically testing charts, but these experiments were not necessarily published for broader consumption; when the conversation arose in JASA, Croxton had data from years prior ready to provide additional insight. --> This multi-year back-and-forth conversation is interesting in part because it pinpoints an early example of experimental statistical graphics as a discipline of interest to statisticians for both personal investigation and as an audience for the results.

Subsequently, @croxtonBarChartsCircle1927 published additional data assessing the accuracy of estimates from pie and bar charts, varying number of categories, and orientation/alignment of sections.
@croxtonBarChartsCircle1927 found that pie charts matched or surpassed bar charts in most situations, across different scale alignments, and in charts with up to 5 categories (though differences were not statistically significant at 5 categories).
The final paper in this early set of comparisons was @croxtonGraphicComparisonsBars1932, which examined a variety of representations of proportional data, but outside the context of a statistical chart; the experimental stimuli, while applicable to charts, show a focus on general perception (though published in JASA).
Experimental graphics studies have largely gone unreported (if they were conducted) between 1932 and the 1980s, particularly in statistical publications, though the issue of pies and bars occasionally pops up during this period in other disciplines.

# Modern Experimental Graphics

Much later, @clevelandGraphicalPerceptionTheory1984, a statistician and a psychologist by training, respectively, conducted experiments on simple graphical elements.
These experiments were presented as supporting a hierarchy of graphical perception that evolved over the course of the study sequence.
This ranking indicates that aligned length judgments are more accurate than judgments based on area or angle, implying that bar charts are more accurately perceived than pie charts.
While the full hierarchy of feature comprehension is often assumed to be experimentally derived, the experiments presented in the series of Cleveland & McGill papers only inform a few elements of this hierarchy.
As in the late 1920s, the combination of introspection based reasoning and a sequence of simple experiments did not go uncontested.
@spenceDisplayingProportionsPercentages1991 (both psychologists) used forced-choice questions of the form "which is bigger, A or B+C" to investigate the accuracy of pie and bar charts for making part-to-whole comparisons.
The later studies show more of an interdisciplinary conversation around data displays, which have become a general tool not limited to statisticians.
As a result, the literature becomes fragmented; studies are conducted across different disciplines, with different methods, elicitation questions, protocols, and, ultimately, different conclusions.

![Different elicitation methods used to assess accuracy of pie chart estimates.](pie-type-example.png)

@@

@eellsRelativeMeritsCircles1926 study offered significant preliminary insights on the perception of proportions through pie and bar charts; yet, it possesses multiple drawbacks.  While Von Hun has already identified several limits in his paper, further ones warrant discussion.  The sample comprised exclusively psychology students from one institution, hence introducing potential sampling bias and constraining generalisability.  Secondly, the design failed to mitigate the sequence of chart types – participants were exposed to pie charts initially, followed by bar charts three days later, potentially introducing learning or memory effects that could skew the comparison in favour of bar charts.  Third, although participants reported their estimation tactics post hoc (e.g., angle, area), no measures were taken during the task (such as eye tracking or time consumed per chart), which reduces the capacity to analyse the underlying cognitive mechanisms.  Ultimately, the study exclusively assessed whole-number estimates, perhaps obscuring nuanced variations in accuracy or bias among chart styles.

@@

The first experiment investigated how people process pie charts, bar charts, and horizontal divided bar charts by measuring accuracy, processing time, and the effect of the number of components.
Thirty participants were divided into three groups based on task type: simple comparison ($A$ vs. $B$), comparison of a single value with a sum ($A$ vs. $B + C$), and comparison of two sums ($A + B$ vs. $C + D$).
\svp{Were participants estimating ratios, or just judging larger/smaller?}. \mt{Participants were not assessing ratios; they were rendering binary judgements regarding which component (or combination of components) was greater. The experiment employed a two-alternative forced-choice (2AFC) design, in which participants identified the greater of two possibilities (e.g., $A$ vs. $B$, or $A + B$ vs. $C + D$). Instead of conducting numerical estimation or ratio calculation, they compare relative magnitude.}
Accuracy improved with longer display times, reaching near 100% for simple tasks and around 90% for complex ones. Bar charts performed slightly better than pie charts for simple comparisons under time pressure, but differences between chart types were negligible for complex tasks.

The second experiment yielded similar results but employed a different approach.
It introduced a table instead of a horizontal bar chart, fixed the processing time at 3.0 seconds, reduced the number of component levels, and adopted a within-subject design for tasks.
Additionally, it tested the effect of ordering components by magnitude.
The results confirmed the superiority of bar charts over pie charts for accuracy, particularly for simpler tasks and fewer components.
However, the lack of an ordering effect challenges the common belief that ordered components improve perception.
\svp{How many categories were used? It's possible that this is true for small tables but not as true for larger tables.}
\mt{The experiment included two categorical levels: four and seven components. The limited quantity likely facilitated the lack of an ordering effect, since participants could readily compare values without reliance on positional signals. In larger tables with more components, ordering may significantly enhance the accuracy and efficiency of interpretation.}

The third experiment examined how task complexity and display type (pie chart, bar chart, and table) affect performance using a within-subject design, where each subject performed all three task types.
Unlike Experiment 1, which used a between-subjects design, and Experiment 2, which introduced tables and fixed processing time, Experiment 3 retained pie charts, bar charts, and tables while excluding the horizontal divided bar chart.
Processing time was fixed at 3.0 seconds \svp{You can fix stimulus time but not processing time -- processing takes as long as it takes}, and two levels of components (4 and 7) were tested.
Accuracy declined with task complexity, and while pie and bar charts showed similar performance for simpler tasks, pie charts outperformed the others for complex tasks, likely due to easier component identification and combination.
Adjacency influenced bar chart performance but not pie charts, suggesting that perceptual and cognitive factors favor pie charts for complex tasks.

@@

The experiment has several limitations.
The small and homogeneous sample (mainly students) limits the generalizability of the findings.
Short exposure times (1.0--6.0 seconds) may not reflect real-world scenarios where viewers have more time to process information.\mt{Might delete this paragraph}

@@

@@

Which chart works best really depends on the task.
When people are asked to simply compare two values (A vs B), bar charts consistently come out on top---they're faster and more accurate, and participants tend to prefer them for that reason (@spenceDisplayingProportionsPercentages1991; @simkin1987information; @hill2025bar).
When the task shifts to estimating what proportion one part is of the whole ($A / (A+B)$), pie charts do better, likely because they align with how we naturally think about part-whole relationships (@clevelandGraphicalPerceptionTheory1984; @simkin1987information).
\svp{Is this still true when A and B aren't adjacent in e.g. a 3-4 component chart?} \mt{While pie charts have been shown to support accurate part-to-whole estimations in two-part visualizations (e.g., A and B as adjacent slices), it is unclear whether this advantage persists in more complex charts. When pie charts include three or more segments, and the parts being compared (e.g., A and B) are not adjacent, the viewer must shift focus across non-contiguous regions of the circle. This added perceptual effort may diminish the intuitive advantage typically associated with pie charts in simpler tasks. Further studies are needed to assess whether the part-whole encoding remains effective when the target slices are visually separated—especially in 3- or 4-component pie charts where holistic comparison becomes less direct}

>>>>

The debate between pie charts and bar charts has been ongoing for over a century.
<!-- Ok, the debate is not between the charts -- it's between either proponents of either chart (e.g. the people promoting them) or about the use of each chart.  -->
Numerous studies with distinct methodologies and experimental designs have attempted to 
determine which chart is more effective, but no single consensus has emerged, even after 100 years of experimentation.
In this paper, we assess the existing literature and attempt to reconcile experiments with diverging results.

@@

Although William Playfair is frequently acknowledged with popularizing pie and bar charts in the early 1800s, the earliest documented use of bar charts can be traced to the 14th century, when Nicole Oresme employed bar-like representations in The Latitude of Forms.  Consequently, Playfair's contribution is mostly in the formalization and distribution of these charts rather than their creation. By the turn of the 20th century, strong opinions about the effectiveness of different chart types—especially pies versus bars—were already appearing in textbooks and manuals.

For instance, @eellsRelativeMeritsCircles1926 quotes two contemporary resources that discuss pie charts as "not a desirable form of presentation" [@brintonGraphicMethodsPresenting1914] and, more strongly, "an insult to a man's intelligence" [@karsten1923charts].
@brintonGraphicMethodsPresenting1914 even identifies that the "sector method" is common, but if horizontal bar methods were used more frequently, they would be preferred for both speed and accuracy.
@eellsRelativeMeritsCircles1926 also quotes an excerpt from an introductory statistics textbook [@secristIntroductionStatisticalMethods1925], which argues that pie charts require more cognitive effort to use accurately than a comparable stacked bar chart, calling pie diagrams "clumsy and defective".
These insights mirror today's status quo: pie charts are stubbornly common, even though many design guidelines recommend against their use. Building on these early critiques, @wickhamGraphicalCriticismHistorical2013 observed that design guidelines anticipated later empirical findings across many visualization types.

@wickhamGraphicalCriticismHistorical2013 previously noted the prescience of design guidelines prior to empirical studies validating their recommendations across different types of visualizations. [Ok, can you connect this better to the previous paragraph somehow?]{.svp}

@@

Interestingly, the question of pies and bars motivated some of the first empirical evaluations of charts, published in the Journal of the American Statistical Association: @eellsRelativeMeritsCircles1926, @vonhuhnFurtherStudiesGraphic1927, and @croxtonBarChartsCircle1927.
Experimenters approached the problem head-on, directly asking participants to estimate quantities from charts, and evaluated participants on accuracy and, in some cases, speed.
While more sophisticated testing methodologies have been developed over the past century [@vanderplasTestingStatisticalCharts2020], studies are still conducted today using direct estimation approaches [@vanderplasFramedReproducingRevisiting2019].
The phrasing of direct estimation questions is of primary importance: a simple mathematical transformation of the estimated quantity can have an outsized effect on the results. [One important insight from both historical and modern work is that the phrasing of the estimation task itself can strongly influence performance.
For instance, @croxtonBarChartsCircle1927 asked participants to "estimate what percentage each sector represented," whereas @eellsRelativeMeritsCircles1926 phrased his question more generally, asking participants to “judge the relative size of the sectors.” (See Figure 1 for a visual comparison.) These subtle differences in wording can lead to significantly different error rates—underscoring how critical question design is in graphical perception research.]{.mt}

::: {.center}

\begin{figure}[ht]
\centering
\begin{minipage}{0.32\textwidth}
  \includegraphics[width=\linewidth]{eells-1.png}
\end{minipage}
\begin{minipage}{0.32\textwidth}
  \includegraphics[width=\linewidth]{eells-2.png}
\end{minipage}
\begin{minipage}{0.32\textwidth}
  \includegraphics[width=\linewidth]{crox.png}
\end{minipage}
\caption{Sample stimuli from Eells (1926), where participants were asked to estimate proportions directly from unlabeled pie charts.Diagrams from Croxton (1927), using labeled segments to compare accuracy in circle and bar comparisons*}
\end{figure}

:::

[You might want to demonstrate this with an example, perhaps showing the phrasing of Croxton's study vs. Eells? You could even include screenshots from the papers.]{.svp}

The initial studies of the question of pies and bars failed to identify a clear "best chart" across all situations or even to clearly identify in which situations pie charts or bar charts should be preferred.
@eellsRelativeMeritsCircles1926 found that pie charts were read as accurately as bar charts and with comparable speed.
The simplified nature of the experiment attracted critiques from @vonhuhnFurtherStudiesGraphic1927; attached to these criticisms is a short study previously run by Croxton, which examined estimates of the ratio between smaller and larger sectors of two-category pie and bar charts; these contributions will be addressed separately even though they are often cited as a single paper, even within the JASA records.
@vonhuhnFurtherStudiesGraphic1927's paper provides several examples which highlight concerns with the Eells study, and argues that while pie charts may be better in certain simple situations, bar charts are much more extensible and facilitate more advanced comparisons, but does not include any experimental studies.
<!-- Not much has changed today - experimental work is often criticized by those who provide toy examples but no empirical data due to compromises necessary to ensure experimental control.  --> Croxton's early experimental data (attached to @vonhuhnFurtherStudiesGraphic1927) had a larger number of participants than @eellsRelativeMeritsCircles1926, but only tested two-category charts with only two proportions: .25 and .4.
In addition, Croxton's study elicited estimates from participants as $A/B$ comparisons, where $A + B = 1$, while Eells elicited estimates for $A$ and $B$ separately.
<!-- It seems that prior to @eellsRelativeMeritsCircles1926, some researchers were empirically testing charts, but these experiments were not necessarily published for broader consumption; when the conversation arose in JASA, Croxton had data from years prior ready to provide additional insight. --> 
This multi-year back-and-forth conversation is interesting in part because it pinpoints an early example of experimental statistical graphics as a discipline of interest to statisticians for both personal investigation and as an audience for the results.
[It is both an example of early graphical testing and a cautionary tale, as proponents of different viewpoints were primarily talking past each other. While the riposte in these early papers and the strong opinions in contemporary design guidelines are entertaining to read, they do not generally support informed choices about what chart to use in practice.]{.svp}

Subsequently, @croxtonBarChartsCircle1927 published additional data assessing the accuracy of estimates from pie and bar charts, varying number of categories, and orientation/alignment of sections.
@croxtonBarChartsCircle1927 found that pie charts matched or surpassed bar charts in most situations, across different scale alignments, and in charts with up to 5 categories (though differences were not statistically significant at 5 categories).
The final paper in this early set of comparisons was @croxtonGraphicComparisonsBars1932, which examined a variety of representations of proportional data, but outside the context of a statistical chart; the experimental stimuli, while applicable to charts, show a focus on general perception (though published in JASA).
Experimental graphics studies have largely gone unreported (if they were conducted) between 1932 and the 1980s, particularly in statistical publications, though the issue of pies and bars occasionally pops up during this period in other disciplines.
[However, guidance to favor bar charts over pie charts still shows up in statistical literature [@haemerPresentationProblemsSuiting1949;@haemerPseudoThirdDimension1951;@fienbergGraphicalMethodsStatistics1979;@nisselson1978graphic;@wickhamGraphicalCriticismHistorical2013]. It almost seems like approximately once a generation, statisticians think about graphics, decry the state of the field, and call for a better set of empirical guidelines.]{.svp}

# Modern Experimental Graphics
Approximately 55 years after the original sequence of experiments, @clevelandGraphicalPerceptionTheory1984, a statistician and a psychologist by training, respectively, conducted experiments on simple graphical elements.
These experiments were presented as supporting a hierarchy of graphical perception that evolved over the course of the study sequence.
This ranking indicates that aligned length judgments are more accurate than judgments based on area or angle, implying that bar charts are more accurately perceived than pie charts.
While the full hierarchy of feature comprehension is often assumed to be experimentally derived, the experiments presented in the series of Cleveland & McGill papers only inform a few elements of this hierarchy.
As in the late 1920s, the combination of introspection based reasoning and a sequence of simple experiments did not go uncontested.
@spenceDisplayingProportionsPercentages1991 (both psychologists) used forced-choice questions of the form "which is bigger, A or B+C" to investigate the accuracy of pie and bar charts for making part-to-whole comparisons.
The later studies show that the 1980-1990s experiments were more of an interdisciplinary conversation around data displays, [reflecting the fact that statistical graphics had]{.svp} become a [more]{.svp} general [scientific]{.svp} tool [and were no longer]{.svp} limited to statisticians.
As a result, the literature becomes fragmented; studies are conducted across different disciplines, with different methods, elicitation questions, protocols, and, ultimately, different conclusions.

![Different methods used to assess accuracy of pie chart estimates. In addition to the different ways to divide up a pie chart's comparisons, there are also different numerical quantities which can be requested: the ratio of C to A + B and the ratio of C to the whole (A + B + C). These different elicitation methods impact the accuracy of participant responses for each type of chart.](pie-slices.png){fig-alt="Three pie charts, divided up in A/B, (A/B)/C, and A/(B/C)/D slices. The first compares A to B, the second compares A+B to C, and the third compares A+D to B+C. Slices that are grouped (A/B) and (A/D), (B/C) are colored similarly."}

@@

@eellsRelativeMeritsCircles1926 study offered significant preliminary insights on the perception of proportions through pie and bar charts; yet, it possesses multiple drawbacks.  While Von Hun has already identified several limits in his paper, further ones warrant discussion.  The sample comprised exclusively psychology students from one institution, hence introducing potential sampling bias and constraining generalisability.  Secondly, the design failed to mitigate the sequence of chart types – participants were exposed to pie charts initially, followed by bar charts three days later, potentially introducing learning or memory effects that could skew the comparison in favor of bar charts.  Third, although participants reported their estimation tactics post hoc (e.g., angle, area), no measures were taken during the task (such as eye tracking or time consumed per chart), which reduces the capacity to analyse the underlying cognitive mechanisms.  Ultimately, the study exclusively assessed whole-number estimates, perhaps obscuring nuanced variations in accuracy or bias among chart styles.

@@

The first experiment investigated how people process pie charts, bar charts, and horizontal divided bar charts by measuring accuracy, processing time, and the effect of the number of components.
Thirty participants were divided into three groups based on task type: simple comparison ($A$ vs. $B$), comparison of a single value with a sum ($A$ vs. $B + C$), and comparison of two sums ($A + B$ vs. $C + D$).

\svp{Were participants estimating ratios, or just judging larger/smaller?}. \mt{Participants were not assessing ratios; they were rendering binary judgements regarding which component (or combination of components) was greater. The experiment employed a two-alternative forced-choice (2AFC) design, in which participants identified the greater of two possibilities (e.g., $A$ vs. $B$, or $A + B$ vs. $C + D$). Instead of conducting numerical estimation or ratio calculation, they compare relative magnitude.}
Accuracy improved with longer display times, reaching near 100% for simple tasks and around 90% for complex ones. Bar charts performed slightly better than pie charts for simple comparisons under time pressure, but differences between chart types were negligible for complex tasks.

[Were participants estimating ratios, or just judging larger/smaller?]{.svp} Accuracy improved with longer display times, reaching near 100% for simple tasks and around 90% for complex ones.
Bar charts performed slightly better than pie charts for simple comparisons under time pressure, but differences between chart types were negligible for complex tasks.

The second experiment yielded similar results but employed a different approach.
It introduced a table instead of a horizontal bar chart, fixed the processing time at 3.0 seconds, reduced the number of component levels, and adopted a within-subject design for tasks.
Additionally, it tested the effect of ordering components by magnitude.
The results confirmed the superiority of bar charts over pie charts for accuracy, particularly for simpler tasks and fewer components.
However, the lack of an ordering effect challenges the common belief that ordered components improve perception.

\svp{How many categories were used? It's possible that this is true for small tables but not as true for larger tables.}
\mt{The experiment included two categorical levels: four and seven components. The limited quantity likely facilitated the lack of an ordering effect, since participants could readily compare values without reliance on positional signals. In larger tables with more components, ordering may significantly enhance the accuracy and efficiency of interpretation.}

[How many categories were used? It's possible that this is true for small tables but not as true for larger tables.]{.svp}

The third experiment examined how task complexity and display type (pie chart, bar chart, and table) affect performance using a within-subject design, where each subject performed all three task types.
Unlike Experiment 1, which used a between-subjects design, and Experiment 2, which introduced tables and fixed processing time, Experiment 3 retained pie charts, bar charts, and tables while excluding the horizontal divided bar chart.
Processing time was fixed at 3.0 seconds [You can fix stimulus time but not processing time -- processing takes as long as it takes]{.svp}, and two levels of components (4 and 7) were tested.
Accuracy declined with task complexity, and while pie and bar charts showed similar performance for simpler tasks, pie charts outperformed the others for complex tasks, likely due to easier component identification and combination.
Adjacency influenced bar chart performance but not pie charts, suggesting that perceptual and cognitive factors favor pie charts for complex tasks.

@@

The experiment has several limitations.
The small and homogeneous sample (mainly students) limits the generalizability of the findings.

Short exposure times (1.0--6.0 seconds) may not reflect real-world scenarios where viewers have more time to process information.\mt{Might delete this paragraph}

@@

Short exposure times (1.0--6.0 seconds) may not reflect real-world scenarios where viewers have more time to process information.
[Why do psycologists use short exposure times? Look into this -- there's a reason]{.svp} The artificial setting (controlled computer-based tests) fails to capture real-world variability, such as distractions and viewing distance.
Reducing exposure time to 1.0 second in Experiment 4 may have biased results toward pie charts by encouraging intuitive rather than analytical processing

@@

Which chart works best really depends on the task.
When people are asked to simply compare two values (A vs B), bar charts consistently come out on top---they're faster and more accurate, and participants tend to prefer them for that reason (@spenceDisplayingProportionsPercentages1991; @simkin1987information; @hill2025bar).

When the task shifts to estimating what proportion one part is of the whole ($A / (A+B)$), pie charts do better, likely because they align with how we naturally think about part-whole relationships (@clevelandGraphicalPerceptionTheory1984; @simkin1987information).
\svp{Is this still true when A and B aren't adjacent in e.g. a 3-4 component chart?} \mt{While pie charts have been shown to support accurate part-to-whole estimations in two-part visualizations (e.g., A and B as adjacent slices), it is unclear whether this advantage persists in more complex charts. When pie charts include three or more segments, and the parts being compared (e.g., A and B) are not adjacent, the viewer must shift focus across non-contiguous regions of the circle. This added perceptual effort may diminish the intuitive advantage typically associated with pie charts in simpler tasks. Further studies are needed to assess whether the part-whole encoding remains effective when the target slices are visually separated—especially in 3- or 4-component pie charts where holistic comparison becomes less direct}

When the task shifts to estimating what proportion one part is of the whole ($A / (A+B)$), pie charts do better, likely because they align with how we naturally think about part-whole relationships (Cleveland & McGill, 1984; Simkin & Hastie, 1987).
[Is this still true when A and B aren't adjacent in e.g. a 3-4 component chart?]{.svp} [While pie charts have been shown to support accurate part-to-whole estimations in two-part visualizations (e.g., A and B as adjacent slices), it is unclear whether this advantage persists in more complex charts. When pie charts include three or more segments, and the parts being compared (e.g., A and B) are not adjacent, the viewer must shift focus across non-contiguous regions of the circle. This added perceptual effort may diminish the intuitive advantage typically associated with pie charts in simpler tasks. Further studies are needed to assess whether the part-whole encoding remains effective when the target slices are visually separated—especially in 3- or 4-component pie charts where holistic comparison becomes less direct]{.mt}
